Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-27 paper-for-archive/paper.tex
Split graphs into subfigures and fixed minor grammar errors

\documentclass[9pt,preprint,nonatbib]{sigplanconf}
\usepackage{balance}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{etex}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage[pdf]{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
%\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage[numbers,sort&compress]{natbib}
% to save space in lists
\usepackage{enumitem}
\setlist{nolistsep}

@@

\added{While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).   }   \todelete{Designing  a  liveness  based  garbage
  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
For  these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for a
eager language.

@@

\newcommand{\wdh}{.33\textwidth}
\newcommand{\LGCLine}{blue}
\newcommand{\RGCLine}{red}
\newcommand{\ReachLine}{black}
\newcommand{\UseLine}{light-blue}

\begin{figure*}[t!]
  \includegraphics[width=\wdh]{fibheap.pdf}
  \includegraphics[width=\wdh]{sudoku.pdf}
  \includegraphics[width=\wdh]{nperm.pdf}\\
  \includegraphics[width=\wdh]{paraffins.pdf}
  \includegraphics[width=\wdh]{lcss.pdf}
  \includegraphics[width=\wdh]{huffman.pdf}\\
  \includegraphics[width=\wdh]{knightstour.pdf}
  \includegraphics[width=\wdh]{nqueens.pdf}
  \includegraphics[width=\wdh]{deriv.pdf}\\
  \includegraphics[width=\wdh]{treejoin.pdf}
  \includegraphics[width=\wdh]{lambda.pdf}
  \includegraphics[width=\wdh]{gc_bench.pdf}
  \caption{Memory usage.  
The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The \ReachLine\ curve represents the number of
reachable cells and the  \UseLine\ curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage} \figrule
%\vspace*{-1mm}
\end{figure*}

%\vspace*{-6.5mm}
\section{The GC scheme}
\label{sec:GC-scheme}
In        the       liveness        analysis       described        in
Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
variable is  derived from dependence  chains along all paths  from the
program point  where the closure  was created to  reachable evaluation
points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
$\epath_2$ are two such evaluation points.  During GC,
we would like to use a more precise liveness, based on the actual paths
taken  during  execution.   Therefore   we  create  separate  liveness
automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
addition to  automata for dependences  along paths to  both $\epath_1$
and $\epath_2$.  The closure carries  the liveness environment for its
free  variables (as  pointers  to automata,  one  for each  variable).
Initially the liveness  environment is based on  the dependences along
both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
\SIF\ condition, the liveness environments are updated to one based on
either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
collections are based on more precise liveness information.

@@

Memory    usage    graphs    for     the    benchmarks    are    shown
Figure~\ref{fig:memory-usage}.  In  all the  programs we can  see that
the curve  corresponding to  LGC (\LGCLine\ line)  dips below  the RGC
curve (\RGCLine\ line)  during GC.  The graphs also  include the curve
for reachable cells (\ReachLine) and live cells (\UseLine). These were
obtained by forcing  RGC to run at very high  frequency. The curve for
live  cells were  obtained by  recording  heap access  times and  post
processing the data at  the end of the program. Note  that the size of
an LGC cell is 1.16 times the size of a RGC cell.

As  demonstrated by  the gap  between the  \RGCLine\ and  the \UseLine\
lines, a  large number of  cells which are  unused by the  program are
still copied  during RGC. LGC does  a much better job  of closing this
gap but still falls short of the  precision achieved by LGC in case of
eager languages ~\cite{asati14lgc}.  A major source of inefficiency in
LGC is  multiple traversals  of already copied  heap cells.  Since LGC
does not mark the heap cells after the first visit, the same cells can
be  repeatedly  visited  with   different  liveness  states.  We  have
mitigated  this  problem  by   implementing  a  heuristic  which  tags
\CONS\ cells with the set of states already visited.

>>>>

\documentclass[9pt,preprint,nonatbib]{sigplanconf}
\usepackage{balance}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{etex}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx, subcaption}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage[pdf]{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
%\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage[numbers,sort&compress]{natbib}
% to save space in lists
\usepackage{enumitem}
\setlist{nolistsep}

@@

\added{While  the idea  of  using static  analysis  to improve  memory
  utilization  has  been  shown  to   be  effective  for  {\em  eager}
  languages~\cite{asati14lgc,       HofmannJ03,       inoue88analysis,
    lee05static}, a straightforward extension  of the technique is not
  possible  for  lazy  languages,  where  heap-allocated  objects  may
  include   closures    (runtime   representations    of   unevaluated
  expressions).   }   \todelete{Designing  a  liveness  based  garbage
  collector   for  lazy   languages  poses   significant  challenges.}
Firstly, since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.  Moreover, for liveness-based  GC to be effective, we
need to  extend it  to closures  apart from  evaluated data.   Since a
closure can escape the scope in which  it was created, it has to carry
the  liveness  information  of   its  free  variables.   As  execution
progresses  and possible  future uses  are eliminated,  we update  the
liveness  information  in  a  closure with  a  more  precise  version.
For  these  reasons,  the  garbage   collector  also  becomes
significantly more  complicated than a liveness-based  collector for an
eager language.

@@

\newcommand{\wdh}{.75\textwidth}
\newcommand{\LGCLine}{blue}
\newcommand{\RGCLine}{red}
\newcommand{\ReachLine}{black}
\newcommand{\UseLine}{light-blue}

\begin{figure*}[t!]
 \caption{Memory usage.  
The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The \ReachLine\ curve represents the number of
reachable cells and the  \UseLine\ curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage}
 \begin{subfigure}{\textwidth}
   \includegraphics[width=\wdh]{fibheap.pdf}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
   \includegraphics[width=\wdh]{sudoku.pdf}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
   \includegraphics[width=\wdh]{nperm.pdf}
 \end{subfigure}
\end{figure*}

\begin{figure*}[t!]
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{lcss.pdf}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{huffman.pdf}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{knightstour.pdf}
  \end{subfigure}
\end{figure*}

\begin{figure*}[t!]
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{deriv.pdf}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{treejoin.pdf}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{lambda.pdf}
  \end{subfigure}
\end{figure*}

\begin{figure*}[t!]
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{gc_bench.pdf}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\wdh]{nqueens.pdf}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
   \includegraphics[width=\wdh]{paraffins.pdf}
  \end{subfigure} 
 \figrule
\end{figure*}

\section{The GC scheme}
\label{sec:GC-scheme}
In        the       liveness        analysis       described        in
Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
variable is  derived from dependence  chains along all paths  from the
program point  where the closure  was created to  reachable evaluation
points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
$\epath_2$ are two such evaluation points.  During GC,
we would like to use a more precise liveness, based on the actual paths
taken  during  execution.   Therefore   we  create  separate  liveness
automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
addition to  automata for dependences  along paths to  both $\epath_1$
and $\epath_2$.  The closure carries  the liveness environment for its
free  variables (as  pointers  to automata,  one  for each  variable).
Initially the liveness  environment is based on  the dependences along
both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
\SIF\ condition, the liveness environments are updated to one based on
either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
collections are based on more precise liveness information.

@@

Memory   usage    graphs   for    the   benchmarks   are    shown   in
Figure~\ref{fig:memory-usage}.  In  all the  programs we can  see that
the curve  corresponding to  LGC (\LGCLine\ line)  dips below  the RGC
curve (\RGCLine\ line)  during GC.  The graphs also  include the curve
for reachable cells (\ReachLine) and live cells (\UseLine). These were
obtained by forcing  RGC to run at very high  frequency. The curve for
live  cells were  obtained by  recording  heap access  times and  post
processing the data at  the end of the program. Note  that the size of
an LGC  cell is 1.16 times  the size of  a RGC cell as  it potentially
might have  to store references  to liveness DFA of  closure arguments
(if the cell is a closure).

As  demonstrated by  the gap  between the  \RGCLine\ and  the \UseLine\
lines, a  large number of  cells which are  unused by the  program are
still copied  during RGC. LGC does  a much better job  of closing this
gap but still falls short of the  precision achieved by LGC in case of
eager languages ~\cite{asati14lgc}.  A major source of inefficiency in
LGC is  multiple traversals  of already copied  heap cells.  Since LGC
does not mark the heap cells after the first visit, the same cells can
be  repeatedly  visited  with   different  liveness  states.  We  have
mitigated  this  problem  by   implementing  a  heuristic  which  avoids 
revisiting closures and function arguments more than once. 
%% Since the 
%% liveness automata for a closure argument encompasses all possible uses 
%% of the closure, multiple traversals for closures can be safely avoided.
